---
title: "Better Season Results Visualizations"
author: "Philip Bulsink"
date: "August 21, 2016"
output: html_document
status: process
published: true
layout: post
excerpt_separator: <!--more-->
tags: R hockey visualizations predictions
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In the past, we've Having a table of expected positions looks terrible. It's hard to read, doesn't fit in a page, and that much data is hard to absorb on the face of it. 

Luckily, R has some great visualization tools in `ggplot2`. I'll demonstrate some new ways to visuallize results, based on past posts' predictions, and I'll use these in future posts.

<!--more-->

Looking at the data inspired me to match the stacked probability curves used by fivethirtyeight in posts [like these](https://fivethirtyeight.com/features/florida-ohio-democratic-primary-preview/). 

To do this, using ggplot's `facet` feature, I needed to reshape the data. A simiple forloop and lapply gets me what I need:

```{r echo=FALSE}
nhl_2016_predicted_standings<-readRDS("./_data/nhl_2016_predicted_standings.RDS")
```

```{r}
library(reshape2)

nhl2016predictexpand<-list()

for (t in c(1:30)){
    nhl2016predictexpand[[rownames(nhl_2016_predicted_standings)[t]]]<-unlist(lapply(c(1:30), function(y) rep(y, nhl_2016_predicted_standings[t,y])))
}

df.results<-as.data.frame(t(as.data.frame(nhl2016predictexpanded)))
df.results$Team<-rownames(df.results)
df.results.m<-melt(df.results, "Team")
```

This expands the data from summary form we saw to a (very) long form, listing each teams' position the number of times they placed there. Thus, this is a 30,000 (30 teams x 1000 simulations) length frame.

Data in this form allows us to use ggplot's facet:

```{r 2015_prediction_plot}
ggplot(df.results.m, aes(x=value)) + 
    geom_density(alpha=0.3, adjust=1.5) + 
    facet_grid(Team~., scales="free") + 
    theme(strip.text.y = element_text(angle=0), strip.background = element_blank()) + 
    scale_y_continuous(breaks=NULL) +
    xlab("End-Of-Season Position") +
    ylab("Probability") +
    ggtitle("Probability of finishing at each position in 2015")
```

There's obviously lots of work that can be down with these plots (for another day), but it gives a much clearer picture of who expects to finish where.

