---
title: "Predicting 2016-2017 NHL Season Results"
author: "Philip Bulsink"
date: "August 18, 2016"
output: html_document
---

Now to some new posts! 

When making a prediction engine, it's always fun to see what next season looks like. We have the schedule for the 2016-2017 NHL season, and we have all the data from the past seasons, so let's get some calculations going! 

<!--more-->

This will hopefully eventually maybe become a daily-updating model, that watches teams as they play though the season to see what happens. The cool thing would be to have available the effect of different scenarios, such as winning 'tonight's' game, or someone elses' fate depending on your own results. These are maybe eventual projects.

All the code I'm using is available in earlier posts, so all you need to know is:
- For the time dependant Dixon-Coles, I'm using data for the past 10 years only, including playoffs, with a &xi; value of 0.0191464. 
- For the quick Dixon-Coles, I'm using only 2 years of data. 
- Each team has 0.5 chance of winning if a game goes to overtime.
- I'm simulating the season 1,000 times for each model. 

```{r load_functions_data echo=False}
source("./_rscripts/dcDataPrep.R")
source("./_rscripts/fastDC.R")
source("./_rscripts/gameAndSeasonPredicting.R")
source("./_rscripts/timeWeightedDC.R")
source("./_rscripts/plotPedictions.R")

nhl_all<-as.data.frame(readRDS("./_data/hockeyData.Rds"))
nhl_past_ten<-nhlDataPrep(nhl_all[nhl_all$Date>as.Date("2006-07-01"),])
nhl_past_ten<-droplevels(nhl_past_ten)
nhl_past_two<-nhl_past_ten[nhl_past_ten$Date>as.Date("2014-07-01"),]
```

```{r prep_res echo=FALSE}
ten_xi<-0.0191464
res_ten<-doDCPrediction(nhl_past_ten, xi=ten_xi, currentDate=as.Date('2016-10-12'))
fast_m<-doFastFit(nhl_past_two)
res_two<-doFastDC(fast_m, nhl_past_two)
```

```{r predictions echo=FALSE}
future_games<-nhlFutureGames(read.csv('./_data/20162017.csv'))
f_stats<-makeStatsTable(future_games)
f_stats$OT.Win.Percent<-rep(0.5)

nhl_2017_ten<-simulateSeason(res_ten, future_games, past_results=NULL, f_stats, n=1000)
#nhl_2017_two<-simulateSeason(res_two, future_games, past_results=NULL, f_stats, n=1000)

```

```{r prediction2017_ten echo=FALSE}
plotPredictions(nhl_2017_ten, title="Results of Time-Weighted DC Prediction of 2016-2017", n_sims=1000)
```

```{r prediction2017_two, echo=FALSE}
#plotPredictions(nhl_2017_two, title="Results of Fast DC Prediction of 2016-2017", n_sims=1000)
```
