---
title: "NHL Stadium Locations"
author: "Philip Bulsink"
date: '2017-03-10'
output: html_document
status: process
published: true
layout: post
excerpt_separator: <!--more-->
maps: true
tags: R hockey maps arenas
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As part of the Data Science Specialization offered by John Hopkins through Coursera, I have a project to create an 'interactive map' in a rmarkdown post. I figured I should map where the NHL stadiums are.

<!--more-->

We'll start by downloading the locations of all of the stadiums. There's a github with these locations. Although the data is possibly a bit out of date (Edmonton has a new stadium, for example), there is a list of locations available on github. We'll download that, convert from JSON to a data frame, then plot them with leaflet, an R package wrapper of a javascript tool by the same name.

```{r get_data}
library(RCurl)
library(rjson)
url<-'https://raw.githubusercontent.com/nhlscorebot/arenas/master/teams.json'
rawjson<-getURL(url)
locations<-fromJSON(rawjson)
```

The locations aren't laid out in a simple to process (or simple to convert to data frame) format, so we'll clean the data a bit.
```{r clean_data}
arenas<-data.frame('team'=names(locations))
uloc<-unlist(locations)
arenas$name<-uloc[seq(1, length(uloc), by=3)]
arenas$lat<-uloc[seq(2, length(uloc), by=3)]
arenas$lng<-uloc[seq(3, length(uloc), by=3)]
arenas$lat<-as.numeric(arenas$lat)
arenas$lng<-as.numeric(arenas$lng)
arenas$label<-paste0(arenas$team, ' - ', arenas$name)
```

With the data ready, we can build our map. It's quite simple. 
```{r leafletmap}
library(leaflet)
library(webshot)
mymap <- arenas %>%
    leaflet() %>%
    addTiles() %>%
    addMarkers(popup = arenas$label)

mymap
```

I'll note, however, that getting it to show on github pages is not easy. So this is just an iframe of it hosted on RPubs.

<iframe width="700" height="500" src="http://rpubs.com/pbulsink/arenamap">&nbsp;</iframe>
